---
title: "woRkshop 01 Exercise 02"
subtitle: "Basic data import"
author: "David Lovell"
date: "24/08/2020"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls()) # Clear workspace. 
knitr::opts_chunk$set(echo = TRUE)
```

## About this exercise

**Exercise 02** gives you an opportunity to gain experience in

* Importing data from `.csv`, `.txt` and `.xlsx` files
* Using the a range of commands to inspect the data you have imported
* Saving and loading R objects

Importing data into `R` is generally the first step in data analysis. This exercise introduces you to basic import mechanisms for commonly used storage formats

You will need to have the following packages installed to run this exercise
```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
```

You may need to install `tidyverse` or `readxl`:

* You can do this by clicking on the `Packages` tab in `RStudio` then hitting `Install` and typing the name of the package
    * You can also type `install.packages("tidyverse")`, `install.packages("readxl")`
    * **Note for virtual machine users:** If you have `R` installed on a VM, you may also need to 

```
sudo apt-get install libcurl4-openssl-dev
sudo apt-get install libssl-dev
sudo apt-get install libxml2-dev
```


### About the data used in this exercise

[Glassdoor is one of the world’s largest job and recruiting sites](https://www.glassdoor.com/about-us/)

* In 2017, Glassdoor's Chief Economist, Dr Andrew Chamberlain, published [How to Analyze Your Gender Pay Gap: An Employer's Guide](https://www.glassdoor.com/research/how-to-analyze-gender-pay-gap-employers-guide/)
  * There is a copy of this report in the `./pdf` directory: it's a good example of using `R` in a professional setting... well worth a read
* I have used the data from that report downloaded as `data.csv`

### Commands you will need to use in this exercise

Use the `Help` function in `RStudio` to understand more about these

* `read_csv()` is the `tidyverse` command to ingest files of comma separated values (`.csv`).
  * You must specify `file` the path to the file you are reading in
  * This command improves on `R`'s base command `read.csv()`
* `str()` compactly displays the structure of an arbitrary R object
  * This function tells you the structure of what you have read in
  * By default `R` often uses special methods to print different kinds of objects in useful ways
  * However, these print methods can mask information about the true natuure of the objects
* `names()`, `dimnames()`, `nrow()`, `ncol()`

## Exercise 02.01: Read `data.csv`

1. Use `read_csv()` to read in the file `./data/data.csv` into a dataframe called `pay.csv`
1. View `pay.csv` by clicking on its name in the `Environment` pane
    1. Inspect its structure by clicking on the blue arrow button icon at the left of its name
1. Use `str()`, `names()`, `dimnames()`, `nrows()`, `ncol()` to inspect this dataframe

### Solutions and commentary

* Look at how this code chunks is named
  * This allows you to [fold code and navigate to different sections and chunks](https://support.rstudio.com/hc/en-us/articles/200484568-Code-Folding-and-Sections)
```{r read-data.csv}
pay.csv <- read_csv("./data/data.csv")
```

Compactly display the structure of the dataframe with `str()`
```{r inspect-pay.csv}
str(pay.csv)
names(pay.csv)
# I've commented out dimnames() to save paper: it returns all 1000 rownames
# dimnames(pay.csv)
nrow(pay.csv)
ncol(pay.csv)
```

## Exercise 02.02: Read `data.txt` and `data.xlsx`

1. Figure out the structure of the file  `./data/data.txt` and then use a function from the `readr` package to read that file into a dataframe called `pay.txt`. Hints:
    1. Type `help("readr")` in the `Console` to get links to more information about this package
    1. Start typing `readr::` in the `Console` to get a list of possible functions of interest, then hit `F1` for more help
1. Using a similar approach, figure out how to read the file `./data/data.xlsx` into a dataframe called `pay.xlsx`
1. Compare `pay.csv`, `pay.txt` and `pay.xlsx` to see if they contain exactly the same information
    1. `R` has two functions for comparing objects: `identical()` which is very stringent and `all.equal()` which is less



### Solutions and commentary

There are many ways to figure out the structure of a file. Given `./data/data.txt` is a small (probably text) file, you probably would be safe to open it in a text editor. If you have a massive file of unknown content, `od` (octal dump) is a good way to go:
```
$ od -a -N 128 ./data/data.txt
0000000   j   o   b   T   i   t   l   e  ht   g   e   n   d   e   r  ht
0000020   a   g   e  ht   p   e   r   f   E   v   a   l  ht   e   d   u
0000040  ht   d   e   p   t  ht   s   e   n   i   o   r   i   t   y  ht
0000060   b   a   s   e   P   a   y  ht   b   o   n   u   s  cr  nl   G
0000100   r   a   p   h   i   c  sp   D   e   s   i   g   n   e   r  ht
0000120   F   e   m   a   l   e  ht   1   8  ht   5  ht   C   o   l   l
0000140   e   g   e  ht   O   p   e   r   a   t   i   o   n   s  ht   2
0000160  ht   4   2   3   6   3  ht   9   9   3   8  cr  nl   S   o   f
0000200
```
Here we can see that `./data/data.txt` contains horizontal tabs (`ht` characters) and carriage return + new lines (`cr`, `nl`) which suggest it is a tab-delimited text file from a Windows Operating System.

Another approach would be to use `wc -n 100 ./data/data.txt` to sneak a peek at the first 100 characters of the file.


```{r read-data.txt}
pay.txt <- read_tsv("./data/data.txt")
# or
# pay.txt <- pay.read_delim("data/data.txt", "\t")
```

`.xlsx` is the file extension for files from the latest version of Excel

```{r read-data.xlsx}
pay.xlsx <- read_xlsx("./data/data.xlsx")
```

`identical()` tests to see whether objects are _exactly_ identical, down to the last byte. However, objects can differ slightly but still behave the same way in almost all respects: `all.equal()` is the appropriate function to explore that. 

In your data science career you may encounter puzzling situations where you think you have the same data, but subtle differences can lead to different outcomes. Being aware of `identical()`  and `all.equal()` may come in handy.

```{r inspect-pay.txt}
identical(pay.csv, pay.txt)
identical(pay.csv, pay.xlsx)
all.equal(pay.csv, pay.txt)
all.equal(pay.csv, pay.xlsx)
```

## Exercise 02.03: Loading and saving R objects

R provides two file formats of its own for storing data, .RDS and .RData. RDS files can store a single R object, and RData files can store multiple R objects

Let's start with a clean environment
```{r}
rm(list=ls()) # Clear workspace. 
```

1. Create a variable called `x`, then save it to a file using `saveRDS()`
    * Hint: Checkout [Data IO in Hands On Programming with R](https://rstudio-education.github.io/hopr/dataio.html)
2. Remove that variable from your R environment using `rm()` and confirm it is gone using `ls()`
3. Read in the object you just saved to a file using `readRDS()` and assign it to a variable called `y`. Check that you have read the oject you saved
    * Hint: the functions `ls()`, `dir()` and `file.remove()` could come in handy

### Solutions and commentary

Saving objects using `R`'s native format can be useful, especially for precious, large, or complex objects
```{r}
x <- "Save me, save me... save me! I'm restless and I'm far from home - Queen"
saveRDS(x, file = "saved.RDS")  # Save x into a file called "saved.RDS"
rm(x)                           # Remove x from R
ls()                            # Check that it is gone
dir()                           # Check that the file "saved.RDS" is on disk
y <- readRDS("saved.RDS")       # Read "saved.RDS" from disk into y
y                               # Check y's contents
file.remove("saved.RDS")        # Remove "saved.RDS" from disk
ls()
dir()
```





## For further reading

1. Grolemund, G., & Wickham, H. [R for Data Science](https://r4ds.had.co.nz/) Chapter 11: Data import
