---
title: "woRkshop 01 Exercise 03"
subtitle: "Calculating summary statistics"
author: "David Lovell"
date: "24/08/2020"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls()) # Clear workspace. 
knitr::opts_chunk$set(echo = TRUE)
```

## About this exercise

**Exercise 03** gives you an opportunity to gain experience in

* Using base `R` functions to subset, summarise and tabulate data
* Calculating measures of central tendency
* Figuring things out using `R` help

You will need to have the following packages installed to run this exercise
```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
```

### About the data used in this exercise

[Glassdoor is one of the world’s largest job and recruiting sites](https://www.glassdoor.com/about-us/)

* In 2017, Glassdoor's Chief Economist, Dr Andrew Chamberlain, published [How to Analyze Your Gender Pay Gap: An Employer's Guide](https://www.glassdoor.com/research/how-to-analyze-gender-pay-gap-employers-guide/)
    * I have used the data from that report downloaded as `data.csv`

### Commands you will need to use in this exercise

Use the `Help` function in `RStudio` to understand more about these

* `read_csv()` is the `tidyverse` command to ingest files of comma separated values (`.csv`).
  * You must specify `file` the path to the file you are reading in
  * This command improves on `R`'s base command `read.csv()`
* `str()` compactly displays the structure of an arbitrary R object
  * This function tells you the structure of what you have read in
  * By default `R` often uses special methods to print different kinds of objects in useful ways
  * However, these print methods can mask information about the true natuure of the objects
* `names()`, `dimnames()`, `nrows()`, `ncols()`

## Exercise 03.01: Using `summary()` and `table()`

Read in Glassdoor's pay data
```{r read-data.csv}
pay <- read_csv("./data/data.csv")
```

1. Use `summary()` to get a quick summary of the `pay` data
    * If you get stuck use `help("summary")` to figure things out
2. Use `summary()` to summarise only the numeric variables
    * i.e., columns 3, 4, 7, 8, 9
    * i.e., "age", "perfEval", "seniority", "basePay", "bonus"
    * If you have forgotten how to subset data frames by column or name, there's plenty of help from Dr Google. Personally, I recommend [the wisdom of Hadley Wickham](https://adv-r.hadley.nz/subsetting.html#introduction-3)
3. Print out the first 10 rows of the character variables
    * i.e., columns 1, 2, 5, 6
    * i.e., "jobTitle", "gender", "edu", "dept"
4. Use `table()` to tabulate each of the `character` variables of the `pay` data in turn
    * Remember that `my.data$age` is the syntax needed to extract a column of values named `age` from a data frame called `my.data`
5. Use `table()` to give a breakdown of `jobTitle` by `gender`
    * Hint: you can use 
    * `addmargins()` and `proportions()` can be handy too

### Solutions and commentary

The `summary()` function is handy for a quick look 
```{r summary}
summary(pay)
```

If your dataframe has a lot of variables, then you will probably need to subset it. We will see other ways to do this (using `select()` from the `tidyverse` package), but the following approaches are old skool and you should know them
```{r summary-subset1}
summary(pay[,c(3, 4, 7, 8, 9)])
# or
summary(pay[,c("age", "perfEval", "seniority", "basePay", "bonus")])
# or
cols <- c("age", "perfEval", "seniority", "basePay", "bonus")
summary(pay[cols])
```

It's important that you understand how to subset rows and columns. Again, there are prettier ways to do this using functions from the `tidyverse` package, but you should be comfortable with these uglier ways too: they are part of `R`'s heritage and they will serve you well for matrices

```{r summary-subset2}
pay[1:10, c(1, 2, 5, 6)]
pay[1:10, c("jobTitle", "gender", "edu", "dept")]
```

`table()` enables us to count up all the instances of specific values in character variables, or factor variables.  
`with()` is another way to work with (!) data frames that can make for more readable code
```{r table1}
table(pay$jobTitle)
with(pay, table(jobTitle))

# For your interest
with(pay,
     {
       print(table(jobTitle)) # Note that you have to use print()
       print(table(gender))
       print(table(edu))
       table(dept)
     }
)
```

We can get multidimensional tables by specifying multiple variables to tabulate:

```{r table2}
table(pay$jobTitle, pay$gender)
with(pay, addmargins(table(jobTitle, gender)))
with(pay, proportions(table(jobTitle, gender), margin=1)) -> job.gender
round(job.gender,2)
```


## Exercise 03.02: Using `mean()`, `median()` but where's `mode()`?

1. What's the mean bonus in the `pay` data?
2. What's the 10% trimmed mean of these bonuses? (i.e., the mean after removing the highest 10% and lowest 10% of values)
    * Why might you use a trimmed mean?
3. Find the mean of this new variable `age.with.NA` 

```{r age.with.NA}
age.with.NA <- c(NA, pay$age)
```

4. What's the median base pay in the `pay` data?
5. What's the median `age.with.NA`
6. What is/are the most common educational levels in the `pay` data?
    * Remember the discussion of modes in our lecture on Measures of Central Tendency?
    * No? Maybe [stackoverflow](https://stackoverflow.com/questions/2547402/how-to-find-the-statistical-mode) can help?

`help("mean")` is your friend for first five questions. Also, as you are typing functions in the editor, hitting the `tab` key can help you by autocompleting commands or suggesting various options. (Try typing `pay$` then hitting `tab`.)
### Solutions and commentary

Remembering all the options for R functions is hard. In these solutions, try typing `mean(pay$bonus,`  then hitting `tab`

```{r means.and.medians}
mean(pay$bonus)
mean(pay$bonus, trim=0.1)
mean(age.with.NA, na.rm = TRUE)
median(pay$basePay)
median(age.with.NA, na.rm = TRUE)
```

From [stackoverflow](https://stackoverflow.com/questions/2547402/how-to-find-the-statistical-mode) we have a simple (but perhaps not very robust) function to find modes in discrete data
```{r function-Modes}
Modes <- function(x) {
  ux <- unique(x)
  tab <- tabulate(match(x, ux))
  ux[tab == max(tab)]
}
```

Now we can find the mode... or modes of `pay$edu`?
```{r}
Modes(pay$edu)
```
Cool! Let's put on a musical!



## For further reading

1. Grolemund, G., & Wickham, H. (2017). [R for Data Science](https://r4ds.had.co.nz/).
2. Spector, P. (2008). [Data Manipulation with R](https://qut.primo.exlibrisgroup.com/permalink/61QUT_INST/1g7tbfa/alma991009334908704001). Springer New York.

