---
title: "woRkshop 03 Exercise 01"
subtitle: "The Law of Large Numbers"
author: "David Lovell"
date: "9/9/2020"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls()) # Clear workspace. 
knitr::opts_chunk$set(echo = TRUE)
```

## About this exercise

**Exercise 01** gives you an opportunity to gain experience in

* Using simulation to explore the Law of Large Numbers
* Using **R**'s  functions for generating random variates from different probability distributions (see `help("distributions")`)

```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
```

## About random numbers in **R**

**R**'s `stats` package implements functions for the following distributions

| Distribution      | See help for | Type       |
|-------------------|--------------|------------|
| Bernoulli         | `dbinom`     | discrete   |
| beta              | `dbeta`      | continuous |
| binomial          | `dbinom`     | discrete   |
| Cauchy            | `dcauchy`    | continuous |
| chi-squared       | `dchisq`     | continuous |
| exponential       | `dexp`       | continuous |
| F                 | `df`         | continuous |
| gamma             | `dgamma`     | continuous |
| geometric         | `dgeom`      | discrete   |
| hypergeometric    | `dhyper`     | discrete   |
| log-normal        | `dlnorm`     | continuous |
| multinomial       | `dmultinom`  | discrete   |
| negative-binomial | `dnbinom`    | discrete   |
| normal            | `dnorm`      | continuous |
| Poisson           | `dpois`      | discrete   |
| Student's t       | `dt`         | continuous |
| uniform           | `dunif`      | continuous |
| Weibull           | `dweibull`   | continuous |

Each distribution (`---`) has four functions named after it:

* `d---(x,...)` is for "density", the value of
    * the probability density function (PDF) at $x$ if the distribution is continuous
    * the value of the probability mass function (PMF) at $x$, if the distribution is discrete
* `p---(q,...)` is for "probability",
    * the cumulative density function (CDF) at $q$
    * the probability of drawing a value $\leq q$ from the distribution
* `q---(p,...)` is for "quantile"
    * the inverse cumulative density function of $p$
    * the $p^\text{th}$ quantile of the distribution
    * the value that has exactly $p$ of the distribution to the left of it
* `r---(n,...)` is for "random"
    * returns a random sample of size $n$

## Exercise 01.01: Working with random numbers

1. Pick a number... any number, from the normal distribution with mean $6.3$ and variance $4$

Don't forget that `rnorm()` uses standard deviation, not variance, in its arguments:
```{r}
rnorm(n=1, mean=6.3, sd=sqrt(4))
```


2. Do it again. Compare your results

```{r}
rnorm(n=1, mean=6.3, sd=sqrt(4))
```
In comparison to question 1, your result should be _different_.

* How different though?
    * Both should be within $2$ standard deviations either side of $6.3$, i.e., they should be in the range $(2.3,10.3)$ with $99.5\%$ probability.

3. Pick 3 numbers from the exponential distribution that has mean $5$

```{r}
rexp(n=3, rate=1/5)
```


If we want to create a reproducible sequence of random numbers we have to set the seed of the random number generator

```{r}
set.seed(123)
rexp(n=3, rate=1/5)
```


4. I'm sorry, can you pick 6 numbers from that exponential distribution, starting with the same 3 you just gave me?

This is a job for `set.seed()` (see previous question)
```{r}
set.seed(123)
rexp(n=6, rate=1/5)
```

5. In dice, "snake eyes" refers to all dice coming up with 1 spot. There are 5 dice in the game of Yahtzee. What's the probability of snake eyes in Yahtzee? Calculate it directly, and by using `dbinom()`

```{r}
(1/6)^5
dbinom(5,5,1/6)
```

6. What's the probability of throwing exactly three 1s in Yahtzee?


This is the same as asking what is the probability of exactly three successes in five trials where the probability of success in any one trial is $\frac{1}{6}

```{r}
dbinom(3, 5, 1/6)
```

7. What's the median of the standard Cauchy distribution (i.e., with location $0$ and scale $1$)?
  * Calculate the mean of a sample of 1000 random variables from the standard Cauchy distribution
  * Calculate the median of that sample and compare it to the mean.
      * Repeat this process several times and discuss what you see
  
Recall that the median is the $0.5$ quantile of a distribution. Roughly speaking, half the distribution is on one side of the median, half is on the other:
```{r}
qcauchy(0.5)
```

As discussed in lectures, the Cauchy distribution has some pathologies. It's mean and variance are undefined, but the median is a good estimate of its location which, in this case is $0$. Running the following code several times should show that the sample mean jumps around all over the place. Make sure you understand how we can calculate the mean of a sample from a distribution whose mean is undefined.
```{r}
cauchy.1000 <-rcauchy(1000)
data.frame(
  mean=mean(cauchy.1000),
  median=median(cauchy.1000),
  difference=mean(cauchy.1000) - median(cauchy.1000)
  )

```

## Exercise 01.02: The Law of Large numbers (continuous distributions)

The Law of Large Numbers tells us two things (Orloff and Bloom):

* the average of many independent samples is (with high probability) close to the mean of the underlying distribution
* the density histogram of many independent samples is (with high probability) close to the graph of the density of the underlying distribution
    * This is known as Émile Borel's Law of Large Numbers

In this exercise, we are going to explore distributions using the Law of Large numbers and, to do that, we need a way to generate and display large numbers of random numbers from different distributions, along with a means to show their theoretical distibution.

I'll give you the materials

Here's how we can tell `ggplot` to plot a function of our own choosing using `stat_function()` from $x=[-3,3]$. I'm going to choose the probability density function of $N(0, 1)$ the Standard Normal:
```{r}
ggplot() +
  stat_function(
    data=tibble(x=c(-3,3)),
    fun=function(x){dnorm(x, mean=0, sd=1)},
    aes(x=x), color="red"
    )
```

Here's how we can create a `tibble` full of random numbers:
```{r}
N <- 10
tibble(
  x=rnorm(N, mean=0, sd=1)
) -> random.numbers
```

...and here's how we can plot a density, histogram of these numbers:
```{r}
ggplot() +
  geom_histogram(
    data=random.numbers,
    binwidth=0.1,
    aes(x=x, y=..density..)
    )
```


1. Plot the histogram of a large number of samples from the Standard Normal distribution and overlay the theoretical distribution for comparison

First we need to generate a large number of samples
```{r}
N <- 1e5
tibble(
  x=rnorm(N, mean=0, sd=1)
) -> random.numbers
```

...then we combine our calls to `geom_histogram()` and `stat_function()`
```{r}
ggplot() +
  geom_histogram(
    data=random.numbers,
    binwidth=0.1,
    aes(x=x, y=..density..)
    ) +
  stat_function(
    data=tibble(x=c(-3,3)),
    fun=function(x){dnorm(x, mean=0, sd=1)},
    aes(x=x), color="red"
    )
```


2. Pick another one of **R**'s continuous distributions. Read its help file to understand a bit about how it is parameterised, then plot its density histogram and theoretical distribution like you did for the Standard Normal.

  * You will probably have to experiment a bit to find our where the data "lives"
  * I suggest just using `geom_histogram()` to work out the domain of the data
      * Then adding the theoretical distribution as appropriate

I've always wondered what a gamma(5,1) distribution looks like...
```{r}
N <- 1e5
tibble(
  x=rgamma(N, shape=5, rate=1)
) -> random.numbers
```

Let's see where it is
```{r}
p <- ggplot() +
  geom_histogram(
    data=random.numbers,
    binwidth=0.1,
    aes(x=x, y=..density..)
    )
p
```

...and then layer its theoretical distribution on top...
```{r}
p +   stat_function(
  data=tibble(x=c(0,20)),
  fun=function(x){dgamma(x, shape=5, rate=1)},
  aes(x=x), color="red"
)
```


## Exercise 01.03: The Law of Large numbers (discrete distributions)

Do you remember why histograms are inappropriate for discrete data?

The correct way to show discrete probability distributions is to plot their probability mass function (PMF)

Here's how we can do that to display the number of heads we get when we toss 5 fair coins, ten times.

First, let's toss the coins
```{r}
N <- 10
tibble(
  x=rbinom(N, size=5, prob=0.5)
) -> random.numbers
```

Now lets plot the proportion of each different value seen:
```{r}
ggplot(data=random.numbers) +
  geom_bar(aes(x=x, y = ..prop..))
```

I prefer to plot probabilty mass functions using skinny lines, but to do that, we need to tabulate the values that appear and calculate their proportions
```{r}
# Make a nice pipeline function to ...
make.pmf <- . %>%
  table() %>%         # Tabulate the values by name
  as_tibble() %>%     # Convert the table to a tibble so we can...
  rename(x='.') %>%   # Rename the first column then...
  mutate(
    x=as.numeric(x),  # Convert the value names to numbers
    y=n/sum(n)        # Work out their proportion of the total
  )

# Now run it
random.numbers %>% make.pmf() -> empirical.pmf 
```

Now we can plot these proportions using skinny lines
```{r}
p <- ggplot(data=empirical.pmf, aes(x=x, y=y)) +
  geom_segment(aes(xend=x), yend=0) +
  expand_limits(y=0) +
  ylab("probability mass")
p
```

We can now calculate the probability mass of the theoretical distribution
```{r}
tibble(
  x=0:5,
  y=dbinom(x, size=5, prob=0.5)
) -> theoretical.pmf
```

...and add it to our plot

```{r}
p + geom_point(data=theoretical.pmf, color="red")
```

1. Plot the histogram of a large number of samples from the distribution of number of heads in 5 coin tosses and overlay the theoretical distribution for comparison
First make the empirical pmf
```{r}
N <- 1000
tibble(
  x=rbinom(N, size=5, prob=0.5)
) %>% make.pmf() -> empirical.pmf
```

Next, make the theoretical pmf
```{r}
tibble(
  x=0:5,
  y=dbinom(x, size=5, prob=0.5)
) -> theoretical.pmf

```

Then plot
```{r}

ggplot(data=empirical.pmf, aes(x=x, y=y)) +
  geom_segment(aes(xend=x), yend=0) +
  geom_point(data=theoretical.pmf, color="red") +
  expand_limits(y=0) +
  ylab("probability mass")
```

2. Pick another one of **R**'s discrete distributions. Read its help file to understand a bit about how it is parameterised, then plot its empirical and theoretical probability mass function like you did for the binomial.
    * Again, you might have to plot the empirical PMF to gauge where your data are

First make the empirical pmf
```{r}
N <- 1000
tibble(
  x=rhyper(N, m=10, n=20, k=12)
) %>% make.pmf() -> empirical.pmf
```

Next, make the theoretical pmf
```{r}
tibble(
  x=0:12,
  y=dhyper(x, m=10, n=20, k=12)
) -> theoretical.pmf

```

Then plot...
```{r}
ggplot(data=empirical.pmf, aes(x=x, y=y)) +
  geom_segment(aes(xend=x), yend=0) +
  geom_point(data=theoretical.pmf, color="red") +
  expand_limits(y=0) +
  scale_x_continuous(breaks=0:12) +
  ylab("probability mass")

```

...note the use of explicit breaks on the `x` axis.

There can be some really tiny probabilities in theoretical distributions. Let's put the theoretical PMF on a log scale to see that better:
```{r}
ggplot(data=empirical.pmf, aes(x=x, y=y)) +
  geom_point() +
  geom_point(data=theoretical.pmf, color="red") +
  expand_limits(y=0) +
  scale_x_continuous(breaks=0:12) +
  scale_y_log10() +
  ylab("probability mass") 
```



## Handy references

1. Jeremy Orloff and Jonathan Bloom (2018) [Central Limit Theorem and the Law of Large Numbers](https://ocw.mit.edu/courses/mathematics/18-05-introduction-to-probability-and-statistics-spring-2014/readings/MIT18_05S14_Reading6b.pdf )
