---
title: "woRkshop 02 Exercise 01"
subtitle: "Visualising Distributions"
author: "David Lovell"
date: "24/08/2020"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls()) # Clear workspace. 
knitr::opts_chunk$set(echo = TRUE)
```

## About this exercise

**Exercise 01** gives you an opportunity to gain experience in

* Plotting continuous univariate distributions
* Visually comparing continuous univariate distributions
* Plotting discrete univariate distributions
* Plotting bivariate continuous distributions
* Understanding how much pie charts are loathed by most data scientists
* Comparing continuous and empirical distributions

```{r libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
```

You may need to install `tidyverse` or `readxl`:

* You can do this by clicking on the `Packages` tab in `RStudio` then hitting `Install` and typing the name of the package
    * You can also type `install.packages("tidyverse")`, `install.packages("readxl")`
    * **Note for virtual machine users:** If you have `R` installed on a VM, you may also need to 

```
sudo apt-get install libssl-dev
sudo apt-get install libxml2-dev
sudo apt-get install libcurl4-openssl-dev
```

## Exercise 01.01: Plotting continuous univariate distributions

Read the Glassdoor data into `pay`
```{r read-data.csv}
# Note that using  col_types = cols() suppresses messages from read_csv
# https://stackoverflow.com/questions/55687401/suppress-all-messages-warnings-with-readr-read-csv-function
pay <- read_csv("./data/data.csv", col_types = cols())
```

Recall from the Lecture and [R for Data Science, Section 3.2.3](https://r4ds.had.co.nz/data-visualisation.html) that the basic syntax of `ggplot` is
```{r eval=FALSE}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

1. Make a `ggplot` that uses  `pay`, maps the variable `basePay` to the x-coordinate and renders the values as a histogram
    * Hint: you will need to use `geom_histogram()`
2. Experiment with the number of `bins` or `binwidth` to get a sense of how the number of bins can affect the histogram produced, and while you are at it, make sure that your plot includes 0 and make sure you understand the issue with including 0
    * Hint: `help("expand_limits")`
    * See:
        * [To cut or not to cut (the y-axis)](https://www.data-to-viz.com/caveat/cut_y_axis.html)
        * [Misleading axes on graphs](https://www.callingbullshit.org/tools/tools_misleading_axes.html)
        * [It’s OK not to start your y-axis at zero](https://qz.com/418083/its-ok-not-to-start-your-y-axis-at-zero/)
        * [Where to Start and End Your Y-Axis Scale](http://stephanieevergreen.com/y-axis/)
3. Histograms report counts of the number of values that fall into a particular bin. Sometimes, we want to see what _proportion_ of values falls into a bin, such as when we want to overlay a histogram and a kernel density estimate.
    * Look up the [Cookbook for R](http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/) and figure out how to produce a _density_ histogram of `basePay` overlayed with a transparent density plot
    * It's not just histograms that can get bumpier. Add in `adjust = 0.2` to adjust the smoothness of your kernel density estimate.
    * Provide arguments to `labs(title= , subtitle= , caption= , tag= )` so you learn how to label your ggplots

### Solutions and commentary

Note that you can map `basePay` to the x-coordinate in the call to `geom_histogram()` (the histogram layer)
```{r}
ggplot(data=pay) +
  geom_histogram(aes(x=basePay))
```

or in the call to `ggplot()` (in which case all geoms will inherit that aesthetic unless specifically overridden)
```{r}
ggplot(data=pay, aes(x=basePay)) +
  geom_histogram()
```

or anywhere in the chain of `+` operators
```{r}
ggplot(data=pay) + 
  aes(x=basePay) +
  geom_histogram()
```

(If you are interested in why `ggplot2` exists and uses the `+` operator to compose functions see [Hadley Wickham's explanation on Reddit](https://www.reddit.com/r/dataisbeautiful/comments/3mp9r7/im_hadley_wickham_chief_scientist_at_rstudio_and/cvi19ly/).)

You can adjust `binwidth`
```{r}
ggplot(data=pay, aes(x=basePay)) +
  geom_histogram(binwidth=2000) + expand_limits(x = 0)
```

You can adjust `bins`
```{r}
ggplot(data=pay, aes(x=basePay)) +
  geom_histogram(bins=100) + expand_limits(x = 0)
```

You can even adjust `bins` _and_ `binwidth`
```{r}
ggplot(data=pay, aes(x=basePay)) +
  geom_histogram(bins=100, binwidth=2000) + expand_limits(x = 0)
```

I didn't know this. I thought things were like that old joke: "You can pick your friends... you can pick your nose... but you can't pick your friend's nose"

The [Cookbook for R](http://www.cookbook-r.com/Graphs/Plotting_distributions_(ggplot2)/) is an excellent resource that helps us whip up plots like this:

```{r}
bandwidth.adjust <- 0.2
ggplot(data=pay, aes(x=basePay)) +
  # Histogram with density instead of count on y-axis
  geom_histogram(
    aes(y=..density..),
    binwidth=2000, colour="black", fill="white"
    ) +
  geom_density(alpha=.2, fill="#FF6666", adjust = bandwidth.adjust) + 
  expand_limits(x = 0) +
  labs(title="Distribution of basePay",
       subtitle="As a density histogram and kernel density estimate" ,
       caption=sprintf("Bandwidth adjustment = %0.1f", bandwidth.adjust),
       tag="Fig. 1")

```

Note the use of `sprintf()` (**f**ormatted **print**ing to a **s**tring) to print the value of a variable in your title.


## Exercise 01.02: Comparing continuous univariate distributions

1. Starting with `ggplot(data=pay, aes(x=basePay))` add a density plot to visualise the distribution of `basePay`
    * Now map `gender` to the aesthetic of `color` to compare the density
    * Use `geom_rug()` to give a sense of the amount of data underpinning these plots
    * Move the legend to the top right hand corner of the plot with advice from the [Cookbook for R](http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/#changing-the-position-of-the-legend)
2. Now try mapping the `gender` to the y-coordinate aesthetic and render the data using points
    * What's the problem with this visualisation?
    * In your own time, try out the strategies used in the Lecture to address this problem
3. Now using `ggplot(data=pay, aes(x=basePay, y=gender, color=gender))` render the values using boxplots
    * Now use `facet_grid(dept ~ .)` to break the data into small multiples where each row plots the data from a particular department
    * Now facet the data further, breaking up each column by education level
        * What does this enable you to explore?

### Solutions and commentary

First we draw a density plot of `basePay`. I suggest mapping `basePay` to the x-axis in the call to ggplot so that all subsequent geoms inherit that mapping

```{r}
ggplot(data=pay, aes(x=basePay)) +
  geom_density()
```

Next map `gender` to the color aesthetic, again in the call to `ggplot`, throw in a rug for good measure and set the theme to shift the non-data ink of the legend.

```{r}
ggplot(data=pay, aes(x=basePay, color=gender)) +
  geom_density() +
  geom_rug() +
  theme(legend.justification=c(1,1), legend.position=c(1,1))
```

We can do that with points
```{r}
ggplot(data=pay, aes(x=basePay, y=gender)) +
  geom_point()
```

But the problem here is _overplotting_ and, as we showed in the lecture, we can address that using jittering
```{r}
ggplot(data=pay, aes(x=basePay, y=gender)) +
  geom_jitter(height = 0.1)
```

...transparency
```{r}
ggplot(data=pay, aes(x=basePay, y=gender)) +
  geom_jitter(height = 0.05, alpha=0.2)
```

...boxplots
```{r}
ggplot(data=pay, aes(x=basePay, y=gender)) +
  geom_boxplot()
```

...violin plots
```{r}
ggplot(data=pay, aes(x=basePay, y=gender)) +
  geom_violin()
```

...the whole shebang!
```{r}
ggplot(data=pay, aes(x=basePay, y=gender)) +
  geom_violin() +
  geom_boxplot() +
  geom_jitter(height = 0.05, alpha=0.2)
```

If we want to introduce other categorical variables, we can use faceting for one...
```{r}
ggplot(data=pay, aes(x=basePay, y=gender, color=gender)) +
  geom_boxplot() +
  facet_grid(dept ~ .)
```

...or two variables.
```{r}
ggplot(data=pay, aes(x=basePay, y=gender, color=gender)) +
  geom_boxplot() +
  facet_grid(dept ~ edu)
```

Can you see any combinations of Department and Education level that buck the `basePay`-`gender` trend?


## Exercise 01.03: Plotting discrete univariate distributions

This exercise shows how to plot the distribution of discrete variables using a barplot, organise the bars in a useful way that is easy on the eye, then pick out a variable of interest. The motivating problem is to figure out how many Data Scientists are in the organisation in comparison to other jobs

1. Starting with `ggplot(data=pay, aes(x=jobTitle))` add a barplot
2. Use `coord_flip()` to flip the axes for easier reading
    * What order are the bars in?
3. Look at the [`forcats` package](https://forcats.tidyverse.org/articles/forcats.html) introduction and figure out how to order the bars by their frequency
4. Set the fill color aesthetic of the bars to pick out your tribe
    * Hint `ifelse(jobTitle=="Data Scientist", "My tribe", "Others")` creates a variable that is equal to "My tribe" if `jobTitle=="Data Scientist"` else it is "Others"
5. Tidy up the presentation by
    * Using `xlab()` or `labs()` to set the x-axis label (remember the coordinates are flipped)
    * Modifying the name of the discrete _color_ scale to something more meaningful

### Solutions and commentary

The key elements are

3. Changing `aes(x=jobTitle)` to `aes(x=fct_infreq(jobTitle))` to reorder the factor levels of `jobTitle`
4. Picking out a bar of interest by setting the fill aesthetic `aes(fill=ifelse(jobTitle=="Data Scientist", "My tribe", "Others"))`
5. Fixing up the labels and legends with `xlab("Job Title")` and `scale_fill_discrete(name="Tribe")` 

```{r}
ggplot(data=pay, aes(x=fct_infreq(jobTitle))) +
  geom_bar(aes(fill=ifelse(jobTitle=="Data Scientist", "My tribe", "Others"))) +
  xlab("Job Title") +
  scale_fill_discrete(name="Tribe") +
  coord_flip()
```


## Exercise 01.04: Plotting discrete univariate distributions with pie charts

1. Read the help file on `geom_pie()`
    * Hint: there is no `geom_pie()` in `ggplot`
    * Read Stephen Few's [Save the Pies for Desert](https://www.perceptualedge.com/articles/visual_business_intelligence/save_the_pies_for_dessert.pdf) to understand why

## Exercise 01.05: Plotting continuous bivariate distributions

The motivation for this exploration is the question: is there any structure to the relationship between base pay and bonus pay?

1. Using the `pay` data, map `basePay` to the x-axis, `bonus` to the y-axis, and plot the pairs of values as points
    * What do you observe?
2. Put the x and y axis on equal scales with `coord_equal()` and use `expand_limits()` to include zeros
    * What do you observe?
3. Use the `facet_grid(dept ~ edu)` to see if subsets of the data show appreciable structure
    * Suggestion: try this without `coord_equal()`
    * Suggestion: try `geom_density2d()`

### Solutions and commentary

Over the entire datasets `basePay` and `bonus` aren't showing any discernable relationships
```{r}
ggplot(data=pay, aes(x=basePay, y=bonus)) + 
  geom_point()
```

...but beware... that previous graph gives the impression that `basePay` and `bonus` are similarly spread. We need to put the two dollar amounts on the same coordinate scales to appreciate that their have different levels of variation
```{r}
ggplot(data=pay, aes(x=basePay, y=bonus)) + 
  geom_point() +
  expand_limits(x=0, y=0) +
  coord_equal()
```

Let's break things up by department and educational level
```{r}
ggplot(data=pay, aes(x=basePay, y=bonus)) + 
  geom_point() +
  facet_grid(dept ~ edu)
```

...and add a density estimate
```{r}
ggplot(data=pay, aes(x=basePay, y=bonus)) + 
  geom_point() +
  geom_density2d() +
  facet_grid(dept ~ edu)
```

...hmmm. Still no discernable structure. Makes me curious to look elsewhere... like `age`... maybe later.


## Exercise 01.06: Comparing continuous and empirical distributions

After that last exercise, and knowing that this is a simulated dataset, I'm wondering if we can "reverse engineer" how the data was made up. Let's use a QQ-plot to compare the distribution of `basePay` and `bonus` to that of the Standard Normal

1. use a QQ-plot to compare the distribution of `basePay` to that of the Standard Normal
    * Hint: use `geom_qq()` and `geom_qq_line()`
    * Hint: these geoms require the variable of interest to be mapped to the `sample` aesthetic
    * What do you observe?
    
2. Here's a present for you: a bit of code that
    * Creates a function that can return the density of a Normal distribution (`dnorm()`) with the same mean and standard deviation as `basePay`
    * Plots that on top of a kernel density estimate of  `basePay`
    * Run it!

```{r eval=FALSE}
normal <- function(x){dnorm(x, mean=mean(pay$basePay), sd=sd(pay$basePay))}
ggplot(data=pay, aes(x=basePay)) + 
  geom_density() +
  stat_function(fun=normal, aes(color="Normal"), n=201) +
  geom_rug()
```

3. Your challenge: conduct steps 1 and 2 with `bonus`
    * For bonus points and a pick of the box: is `bonus` _mesokurtic_ (Normal), _leptokurtic_ (fat-tailed) or _platykurtic_ (thin-tailed)?
    * Fact: this is the first time I have ever used these words in my professional career

### Solutions and commentary

Over the entire dataset, the quantiles of `basePay` show a pretty linear relationship with the quantiles of the Standard Normal:
```{r}
ggplot(data=pay, aes(sample=basePay)) + 
  geom_qq(alpha=0.1) +
  geom_qq_line()
```

...and when we plot a Normal distribution with the same mean and standard deviation as `basePay` over the empirical kernel density estimate, they fit alright
```{r}
normal <- function(x){dnorm(x, mean=mean(pay$basePay), sd=sd(pay$basePay))}
ggplot(data=pay, aes(x=basePay)) + 
  geom_density() +
  stat_function(fun=normal, aes(color="Normal"), n=201) +
  geom_rug()
```

...but remember from Exercise 01.02 that
```{r}
ggplot(data=pay, aes(x=basePay, color=gender)) +
  geom_density() +
  geom_rug() +
  theme(legend.justification=c(1,1), legend.position=c(1,1))
```

...so `basePay` could be a mix of two quite similar normally distibuted variables.

`bonus` is a different story though...
```{r}
ggplot(data=pay, aes(sample=bonus)) + 
  geom_qq(alpha=0.1) +
  geom_qq_line()
```

...exhibiting a thin-tailed ( _platykurtic_ ) distribution
```{r}
normal <- function(x){dnorm(x, mean=mean(pay$bonus), sd=sd(pay$bonus))}
ggplot(data=pay, aes(x=bonus)) + 
  geom_density() +
  stat_function(fun=normal, aes(color="Normal"), n=201) +
  geom_rug()
```

(I think I would call the distribution of `bonus` "lumpokurtic".)





## For further reading

1. Grolemund, G., & Wickham, H. [R for Data Science](https://r4ds.had.co.nz/) 
1. Wickham, H. [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/)
